\documentclass[10pt, a4paper, twoside]{basestyle}
\usepackage[Mathematics]{semtex}

%%%% Shorthands.

%%%% Title and authors.

\title{Documentation for the symplectic methods}
\date{\printdate{2015-06-06}}
\author{Robin~Leroy (eggrobin)}
\begin{document}
\maketitle
This document expands on the comments at the beginning of\\
\texttt{integrators/symplectic\_runge\_kutta\_nyström\_integrator.hpp}.

\section{Differential equations.}
Recall that the equations solved by this class are
\begin{align}
\tuple{\vq,\vp}\der &=
\vX\of{\vq, \vp, t} = \vA\of{\vq, \vp} + \vB\of{\vq, \vp, t}
\quad\parbox{.4\linewidth}{with $\exp h\vA$ and $\exp h\vB$ known and
$\commutator{\vB}{\commutator{\vB}{\commutator{\vB}{\vA}}}=\nullvec$;}
\label{general}\\
\span\parbox{.7\linewidth}{the above equation, with $\exp h\vA = \Identity+h\vA$,
$\exp h\vB = \Identity+h\vB$,
and $\vA$ and $\vB$ known;}
\label{linear}\\
\vq\dder &= -\matM^{-1} \grad_\vq V\of{\vq, t}\text. \label{rkn}
\end{align}

\section{Relation to Hamiltonian mechanics.}
The third equation above is a reformulation of Hamilton's
equations with a Hamiltonian of the form
\begin{equation}
H\of{\vq,\vp,t} = \frac{1}{2}\Transpose{\vp}\matM^{-1}\vp + V\of{\vq, t}\text,
\end{equation}
where $\vp = \matM\vq\der$.

\section{A remark on non-autonomy.}
Most treatments of these integrators write these differential equations as well
as the corresponding Hamiltonian in an autonomous version, thus
$\vX = \vA(\vq, \vp) + \vB(\vq, \vp)$ and
$H\of{\vq,\vp,t} = \frac{1}{2}\Transpose{\vp}\matM^{-1}\vp + V\of{\vq}$.
It is however possible to incorporate time, by considering it as an
additional variable:\[
\tuple{\vq,\vp,t}\der =
\vX\of{\vq, \vp, t} =
\tuple{\vA\of{\vq, \vp}, 1} +
\tuple{\vB\of{\vq, \vp, t}, 0}\text.\]

For equations of the form (\ref{rkn}) it remains to be shown that Hamilton's
equations with quadratic kinetic energy and a time-dependent potential satisfy
$\commutator{\vB}{\commutator{\vB}{\commutator{\vB}{\vA}}}=\nullvec$.
We introduce $t$ and its conjugate momentum $\gcp$ to the phase space,
and write
\[
\tilde{\vq}=\tuple{\vq, t}\text,\quad
\tilde{\vp}=\tuple{\vp, \gcp}\text,\quad
L\of{\tilde{\vp}} = \frac{1}{2}\Transpose{\vp}\matM^{-1}\vp + \gcp\text.
\]
(\ref{rkn}) follows from Hamilton's equations with\[
H\of{\tilde{\vq},\tilde{\vp}} =
L\of{\tilde{\vp}} + V\of{\tilde{\vq}} =
\frac{1}{2}\Transpose{\vp}\matM^{-1}\vp + \gcp + V\of{\vq, t}
\]		
since we then get $t\der = 1$.		
The desired property follows from the following lemma:
\begin{lemma}
Let $L\of{\tilde{\vq},\tilde{\vp}}$ be a quadratic polynomial in $\tilde{\vp}$,
$V\of{\tilde{\vq}}$ a smooth function, $\vA=\Poisson\placeholder L$, and
$\vB=\Poisson\placeholder V$.
Then\[
\commutator{\vB}{\commutator{\vB}{\commutator{\vB}{\vA}}}=\nullvec\text.\]
\end{lemma}
\begin{proof}
It suffices to show that $\Poisson V{\Poisson V{\Poisson L V}} = 0$. It is
immediate that every term in that expression will contain a third order
partial derivative in the $\tilde p_i$ of $L$, and since $L$ is quadratic
in $\tilde{\vp}$ all such derivatives vanish.
\end{proof}
See \cite[p.~26]{McLachlanQuispel2006} for a detailed treatment
of non-autonomous Hamiltonians using an extended phase space.
See \cite[p.~8]{McLachlan1993} for a proof that
$\Poisson V{\Poisson V{\Poisson L V}} = 0$ for arbitrary Poisson tensors.
\section{Composition and first-same-as-last property}
Recall from the comments that each step is computed as
\begin{align*}
\tuple{\vq_{n+1}, \vp_{n+1}} &=
\exp a_{r-1}h\vA \exp b_{r-1}h\vB \dotsb \exp a_0h\vA \exp b_0h\vB 
\tuple{\vq_n, \vp_n}
\text,
\intertext{thus, when $b_0$ vanishes (type $ABA$) or when $a_{r-1}$ does
(type $BAB$),}
\tuple{\vq_{n+1}, \vp_{n+1}} &=
\exp a_{r-1}h\vA \exp b_{r-1}h\vB \dotsb \exp b_1h\vB \exp a_0h\vA
\tuple{\vq_n, \vp_n}\text{, respectively}\\
\tuple{\vq_{n+1}, \vp_{n+1}} &=
\exp b_{r-1}h\vB \exp a_{r-2}h\vA \dotsb \exp a_0h\vA \exp b_0h\vB
\tuple{\vq_n, \vp_n}\text.
\end{align*}
This leads to performance savings.

Let us consider a method of type $BAB$. 
Evidently, the evaluation of $\exp a_0h\vA$
is not required, thus only $r-1$ evaluations of  $\exp \increment t\vA$ are
required.
Furthermore, if output is not needed at step $n$, the computation of
the $\pa{n-1}$th step requires only $r-1$ evaluations of
$\exp \increment t\vB$, since the consecutive evaluations of $\exp b_0h\vB$
and $\exp b_rh\vB$ can be merged by the group property,\[
\exp b_0h\vB\exp b_rh\vB=\exp \pa{b_0+b_r}h\vB\text.
\]

If the equation is of the form \ref{linear}, the latter saving can be achieved even
for dense output, since only one evaluation of $\vB$ is needed to compute the
increments $b_rh\vB$ and $b_0\vB$.

The same arguments apply to type $ABA$.
This motivates the name of the template
parameter \texttt{evaluations}, equal to $r-1$ for methods of type $ABA$ and
$BAB$, and $r$ otherwise.
\end{document}
